Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-08-24 Manuscript/MPESubmission/Revision/TEM_revised_manuscript.tex
taxon -> taxa + aknowledgements

We created matrices using all parameter combinations resulting in 125 ($5^3$) ``missing-data'' matrices.
Note that one of these combinations ($M_L$=0\%; $M_F$=0\% and $N_C$=100) has no missing data so is equivalent to the ``complete'' matrix, thus we have one effectively complete matrix in our 125 ``missing-data'' matrices.
In practice, we first removed the data following the two missing data parameters $M_L$ and $M_F$ and then removed data following the $N_C$ parameters.
To avoid avoid matrices containing taxa without any data (morphological or molecular), we repeated the random deletion until the matrices contained at least 5\% of data for any taxon.
Note that the living taxa always had at least 90\% of data (the 1000 molecular characters).

@@

The effect of the missing data in the fossil record ($M_{F}$) is less than the effect of the $M_{L}$ parameter on clade conservation (Normalised Robinson-Foulds metric) but greater on the displacement of wildcard taxa (Normalised Triplets metric; Fig. 4 and Fig. 5).
This is related to the fact that the Bayesian consensus tree is built using a majority consensus rule.
When the fossil taxa have less data (e.g. $M_{F}$ = 75\%) they will tend to branch with any taxon in the clade that shares most characters with the fossils.
Therefore a majority consensus position is unlikely to exist (i.e. every branching position is represented in $<$ 50\% of the trees in the Bayesian posterior distribution) and the fossil taxa will form a polytomy at the base of the clade.
In this case, the Normalised Robinson-Foulds metric will decrease when the fossil is present near the tips but affects the clade conservation less when fossils are near the root.
Conversely, because a fossil in a high taxonomic level clade has many chances to branch on different nodes within the clade, it will be more likely to act as a wildcard taxon and decrease the Normalised Triplets metric.
Therefore, the $M_{F}$ parameter is likely to affect the Normalised Robinson-Foulds metric less than the Normalised Triplets metric for the Bayesian consensus trees.
Conversely, the same scenario in a Maximum Likelihood framework will lead to a dichotomous branching of the fossils but with low bootstrap support ($<$ 50).
In other words, the Bayesian consensus tree allows a fossil taxon with few data to be placed with a higher confidence at a lower taxonomic level than the Maximum Likelihood tree, where the fossil will be placed with lower confidence at a higher taxonomic level.
We argue that using the Bayesian consensus tree topology is preferable because it is more conservative \citep[e.g.][]{pattinsonphylogeny2014}.

@@

\section{Acknowledgments}
Thanks to Gavin Thomas, Fr\'{e}d\'{e}ric Delsuc, Emmanuel Douzery, Trevor Hodkinson, Andrew Jackson, Nick Matzke, and April Wright for useful comments on our simulation protocol and manuscript. Thanks to Paddy Doyle, Graziano D'Innocenzo and Sean McGrath for assistance with the computer cluster. Thanks to two anonymous reviewers for their useful and encouraging comments. Simulations used the Lonsdale cluster maintained by the Trinity Centre for High Performance Computing and funded through grants from Science Foundation Ireland. This work was funded by a European Commission CORDIS Seventh Framework Programme (FP7) Marie Curie CIG grant (proposal number: 321696).

>>>>

We created matrices using all parameter combinations resulting in 125 ($5^3$) ``missing-data'' matrices.
Note that one of these combinations ($M_L$=0\%; $M_F$=0\% and $N_C$=100) has no missing data so is equivalent to the ``complete'' matrix, thus we have one effectively complete matrix in our 125 ``missing-data'' matrices.
In practice, we first removed the data following the two missing data parameters $M_L$ and $M_F$ and then removed data following the $N_C$ parameters.
To avoid avoid matrices containing taxa without any data (morphological or molecular), we repeated the random deletion until the matrices contained at least 5\% of data for any taxa.
Note that the living taxa always had at least 90\% of data (the 1000 molecular characters).

@@

The effect of the missing data in the fossil record ($M_{F}$) is less than the effect of the $M_{L}$ parameter on clade conservation (Normalised Robinson-Foulds metric) but greater on the displacement of wildcard taxa (Normalised Triplets metric; Fig. 4 and Fig. 5).
This is related to the fact that the Bayesian consensus tree is built using a majority consensus rule.
When the fossil taxa have less data (e.g. $M_{F}$ = 75\%) they will tend to branch with any taxa in the clade that shares most characters with the fossils.
Therefore a majority consensus position is unlikely to exist (i.e. every branching position is represented in $<$ 50\% of the trees in the Bayesian posterior distribution) and the fossil taxa will form a polytomy at the base of the clade.
In this case, the Normalised Robinson-Foulds metric will decrease when the fossil is present near the tips but affects the clade conservation less when fossils are near the root.
Conversely, because a fossil in a high taxonomic level clade has many chances to branch on different nodes within the clade, it will be more likely to act as a wildcard taxon and decrease the Normalised Triplets metric.
Therefore, the $M_{F}$ parameter is likely to affect the Normalised Robinson-Foulds metric less than the Normalised Triplets metric for the Bayesian consensus trees.
Conversely, the same scenario in a Maximum Likelihood framework will lead to a dichotomous branching of the fossils but with low bootstrap support ($<$ 50).
In other words, the Bayesian consensus tree allows a fossil taxon with few data to be placed with a higher confidence at a lower taxonomic level than the Maximum Likelihood tree, where the fossil will be placed with lower confidence at a higher taxonomic level.
We argue that using the Bayesian consensus tree topology is preferable because it is more conservative \citep[e.g.][]{pattinsonphylogeny2014}.

@@

\section{Acknowledgments}
Thanks to Gavin Thomas, Fr\'{e}d\'{e}ric Delsuc, Emmanuel Douzery, Trevor Hodkinson, Andrew Jackson, Nick Matzke, and April Wright for useful comments on our simulation protocol and manuscript. Thanks to Paddy Doyle, Graziano D'Innocenzo and Sean McGrath for assistance with the computer cluster. Thanks to one anonymous reviewer and Alex Pyron for their useful and encouraging comments. Simulations used the Lonsdale cluster maintained by the Trinity Centre for High Performance Computing and funded through grants from Science Foundation Ireland. This work was funded by a European Commission CORDIS Seventh Framework Programme (FP7) Marie Curie CIG grant (proposal number: 321696).
